### 2015-05-06 doc/thesis_puc/evaluation.tex
Fixing typos

We could give precise static types to most of the members
in the \texttt{socket} module, which is the C core.
However, this module includes some functions that
we could not give a precise static type because they
rely on reflection, as the \emph{hard} category shows.
For instance, \texttt{socket.skip} is a function that is
in this category.
We can use this function to choose the number of
values that we want to return.
As an example, calling \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, because \texttt{1} indicates
that we do not want to return the first value.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
As an example, calling \texttt{socket.skip(-1, nil, "hello")}
returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
\texttt{socket.skip} does not skip any values.
As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
Our type system cannot handle the type of negative numbers, as
this requires more complex types such as the refinement types from
hybrid type checking \cite{flanagan2006htc}.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it uses LPeg \cite{lpeg,ierusalimschy2009lpeg}
patterns, and LPeg uses overloaded arithmetic operators to
build LPeg patterns. 
Even though LPeg is the third most popular Lua module,
we cannot not precisely type LPeg patterns because our
type system still does not support overloading arithmetic
operators.
In the \texttt{tllexer} module, we could only give precise static types
to two error reporting functions that it exports.

>>>>

We could give precise static types to most of the members
in the \texttt{socket} module, which is the C core.
However, this module includes some functions that
we could not give a precise static type because they
rely on reflection, as the \emph{hard} category shows.
For instance, \texttt{socket.skip} is a function that is
in this category.
We can use this function to choose the number of
values that we want to return.
As an example, calling \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, because \texttt{1} indicates
that we do not want to return the first value.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
As an example, calling \texttt{socket.skip(-1, nil, "hello")}
returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
\texttt{socket.skip} not skip any values.
As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
Our type system cannot handle the type of negative numbers, as
this requires more complex types such as the refinement types from
hybrid type checking \cite{flanagan2006htc}.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it uses LPeg \cite{lpeg,ierusalimschy2009lpeg}
patterns, and LPeg uses overloaded arithmetic operators to
build LPeg patterns. 
Even though LPeg is the third most popular Lua module,
we cannot precisely type LPeg patterns because our
type system still does not support overloading arithmetic
operators.
In the \texttt{tllexer} module, we could only give precise static types
to two error reporting functions that it exports.

### 2015-05-06 doc/thesis_puc/review.tex
Fixing typos

Abadi and associates \cite{abadi1989dts} extended the simply typed
lambda calculus with the \texttt{Dynamic} type and the \texttt{dynamic}
and \texttt{typecase} constructs, with the aim to safely integrate dynamic
code in statically typed languages.
The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
value and \texttt{T} is the tag that represents the type of \texttt{v}.
The constructs \texttt{dynamic} and \texttt{typecase} are explicit
injection and projection operations, respectively.
That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
Thus, migrating code between dynamic and static type checking requires
changing type annotations and adding or removing \texttt{dynamic} and
\texttt{typecase} constructs throughout the code.

The \emph{quasi-static} type system proposed by Thatte \cite{thatte1990qst}
performs implicit coercions and run-time checks to replace the
\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
Abadi and associates \cite{abadi1989dts}.
To do that, quasi-static typing relies on subtyping with a top type
$\Omega$ that represents the dynamic type, and splits type checking
into two phases.
The first phase inserts implicit coercions from the dynamic type to
the expected type, while the second phase performs what Thatte calls
\emph{plausibility checking}, that is, it rewrites the program to
guarantee that sequences of upcasts and downcasts always have a
common subtype.

@@

Ou and associates \cite{ou2004dtd} specified a language that combines
static types with dependent types.
To ensure safety, the compiler automatically inserts coercions
between dependent code and static code.
The coercions are run-time checks that ensure static code does not
crash dependent code during run-time.

@@

The main goal of gradual typing \cite{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, Siek and Taha \cite{siek2006gradual} extended the simply
typed lambda calculus with the dynamic type $?$, as we can see in
Figure \ref{fig:gtlc}.
In gradual typing, type annotations are optional, and an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
Under these circumstances, we view gradual typing as a way to add
a dynamic type to statically typed languages.

@@

The central idea of gradual typing is the \emph{consistency}
relation, written $T_{1} \sim T_{2}$.
The consistency relation allows implicit conversions to and from the
dynamic type, and disallows conversions between inconsistent types
\cite{siek2006gradual}.
For instance, $\Number \sim \;?$, $? \sim \Number$,
$\String \sim \;?$, and $? \sim \String$,
but $\Number \not\sim \String$, and
$\String \not\sim \Number$.
The consistency relation is both reflexive and symmetric, but
it is neither commutative nor transitive.

@@

Gradual typing \cite{siek2006gradual} is similar to two previous
approaches \cite{abadi1989dts,thatte1990qst}, because they also include
a dynamic type in a statically typed language.
However, these three approaches differ in the way they handle the
dynamic type.
While Siek and Taha \cite{siek2006gradual} rely on the consistency relation,
Abadi and associates \cite{abadi1989dts} rely on type equality with explicit
projections plus injections, and Thatte \cite{thatte1990qst} relies on subtyping.

@@

Herman and associates \cite{herman2007sgt} showed that there is an
efficiency concern regarding the run-time checks, because there are
two ways that casts can lead to unbounded space consumption.
The first affects tail recursion while the second appears when
first-class functions or objects cross the border between
static code and dynamic code, that is, some programs can apply
repeated casts to the same function or object.
Herman and associates \cite{herman2007sgt} use the coercion calculus
outlined by Henglein \cite{henglein1994dts} to express casts
as coercions and solve the problem of space efficiency.
Their approach normalizes an arbitrary sequence of coercions to a
coercion of bounded size.

Another concern about casts is how to improve debugging support,
because a cast application can be delayed and the error related
to that cast application can appear considerable distance
from the real error.
Wadler and Findler \cite{wadler2009wpc} developed \emph{blame calculus}
as a way to handle this issue, and Ahmed and associates \cite{ahmed2011bfa}
extended blame calculus with polymorphism.
Blame calculus is an intermediate language to integrate
static and dynamic typing along with the blame tracking approach
proposed by Findler and Felleisen \cite{findler2002chf}.

On the one hand, blame calculus solves the issue regarding
error reporting;
on the other hand, it has the space efficiency problem reported
by Herman and associates \cite{herman2007sgt}.
Thus, Siek and associates \cite{siek2009casts} extended the coercion
calculus outlined by Herman and associates \cite{herman2007sgt} with
blame tracking to achieve an implementation of the blame calculus that
is space efficient.
After that, Siek and Wadler \cite{siek2010blame} proposed a new solution
that also handles both problems.
This new solution is based on a concept called \emph{threesome},
which is a way to split a cast between two parties into two casts
among three parties.
A cast has a source and a target type (a \emph{twosome}),
so we can split any cast into a downcast from the source to an
intermediate type that is followed by an upcast from the intermediate
type to the target type (a \emph{threesome}).

@@

The indexing of tables appears 130448 times:
86\% of them are for reading a table field while
14\% of them are for writing into a table field.
We can classify the indexing operations that are reads as follows:
89\% of the reads use a literal string key,
4\% of the reads use a literal number key,
and less than 1\% of the reads (10 times) use a literal boolean key.
At all, 93\% of the reads use literals to index a table while
7\% of the reads use non-literal expressions to index a table.
It also worth mentioning that 45\% of the reads are actually
function calls.
More precisely, 25\% of the reads use literals to call a function,
20\% of the reads use literals to call a method, that is,
a function call that uses the colon syntactic sugar, 
and less than 1\% of the reads (195 times) use non-literal expressions
to call a function.
We can also classify the indexing operations that are writes as follows: 
69\% of the writes use a literal string key,
2\% of the writes use a literal number key,
and less than 1\% of the writes (1 time) uses a literal boolean key.
At all, 71\% of the writes use literals to index a table while
29\% of the writes use non-literal expressions to index a table.

We also measured how many files have code that iterate over tables to
observe how frequently iteration is used.
We observed that 23\% of the files have code that iterate over keys
of any value, that is, the call to \texttt{pairs} appears at least
once in these files (the median is twice per file);
21\% of the files have code that iterate over integer keys, that is,
the call to \texttt{ipairs} appears at least once in these files
(the median is also twice per file);
and 10\% of the files have code that use the numeric \texttt{for}
along with the length operator (the median is once per file).

>>>>

Abadi et al. \cite{abadi1989dts} extended the simply typed
lambda calculus with the \texttt{Dynamic} type and the \texttt{dynamic}
and \texttt{typecase} constructs, with the aim to safely integrate dynamic
code in statically typed languages.
The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
value and \texttt{T} is the tag that represents the type of \texttt{v}.
The constructs \texttt{dynamic} and \texttt{typecase} are explicit
injection and projection operations, respectively.
That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
Thus, migrating code between dynamic and static type checking requires
changing type annotations and adding or removing \texttt{dynamic} and
\texttt{typecase} constructs throughout the code.

The \emph{quasi-static} type system proposed by Thatte \cite{thatte1990qst}
performs implicit coercions and run-time checks to replace the
\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
Abadi et al. \cite{abadi1989dts}.
To do that, quasi-static typing relies on subtyping with a top type
$\Omega$ that represents the dynamic type, and splits type checking
into two phases.
The first phase inserts implicit coercions from the dynamic type to
the expected type, while the second phase performs what Thatte calls
\emph{plausibility checking}, that is, it rewrites the program to
guarantee that sequences of upcasts and downcasts always have a
common subtype.

@@

Ou et al. \cite{ou2004dtd} specified a language that combines
static types with dependent types.
To ensure safety, the compiler automatically inserts coercions
between dependent code and static code.
The coercions are run-time checks that ensure static code does not
crash dependent code during run-time.

@@

The main goal of gradual typing \cite{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, Siek and Taha \cite{siek2006gradual} extended the simply
typed lambda calculus with the dynamic type $?$, as we can see in
Figure \ref{fig:gtlc}.
In gradual typing, type annotations are optional, and an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
Under these circumstances, we can view gradual typing as a way to add
a dynamic type to statically typed languages.

@@

The central idea of gradual typing is the \emph{consistency}
relation, written $T_{1} \sim T_{2}$.
The consistency relation allows implicit conversions to and from the
dynamic type, and disallows conversions between inconsistent types
\cite{siek2006gradual}.
For instance, $\Number \sim \;?$, $? \sim \Number$,
$\String \sim \;?$, and $? \sim \String$,
but $\Number \not\sim \String$, and
$\String \not\sim \Number$.
The consistency relation is both reflexive and symmetric, but
it is not transitive.

@@

Gradual typing \cite{siek2006gradual} is similar to two previous
approaches \cite{abadi1989dts,thatte1990qst}, because they also include
a dynamic type in a statically typed language.
However, these three approaches differ in the way they handle the
dynamic type.
While Siek and Taha \cite{siek2006gradual} rely on the consistency relation,
Abadi et al. \cite{abadi1989dts} rely on type equality with explicit
projections plus injections, and Thatte \cite{thatte1990qst} relies on subtyping.

@@

Herman et al. \cite{herman2007sgt} showed that there is an
efficiency concern regarding the run-time checks, because there are
two ways that casts can lead to unbounded space consumption.
The first affects tail recursion while the second appears when
first-class functions or objects cross the border between
static code and dynamic code, that is, some programs can apply
repeated casts to the same function or object.
Herman et al. \cite{herman2007sgt} use the coercion calculus
outlined by Henglein \cite{henglein1994dts} to express casts
as coercions and solve the problem of space efficiency.
Their approach normalizes an arbitrary sequence of coercions to a
coercion of bounded size.

Another concern about casts is how to improve debugging support,
because a cast application can be delayed and the error related
to that cast application can appear considerable distance
from the real error.
Wadler and Findler \cite{wadler2009wpc} developed \emph{blame calculus}
as a way to handle this issue, and Ahmed et al. \cite{ahmed2011bfa}
extended blame calculus with polymorphism.
Blame calculus is an intermediate language to integrate
static and dynamic typing along with the blame tracking approach
proposed by Findler and Felleisen \cite{findler2002chf}.

On the one hand, blame calculus solves the issue regarding
error reporting;
on the other hand, it has the space efficiency problem reported
by Herman et al. \cite{herman2007sgt}.
Thus, Siek et al. \cite{siek2009casts} extended the coercion
calculus outlined by Herman et al. \cite{herman2007sgt} with
blame tracking to achieve an implementation of the blame calculus that
is space efficient.
After that, Siek and Wadler \cite{siek2010blame} proposed a new solution
that also handles both problems.
This new solution is based on a concept called \emph{threesome},
which is a way to split a cast between two parties into two casts
among three parties.
A cast has a source and a target type (a \emph{twosome}),
so we can split any cast into a downcast from the source to an
intermediate type that is followed by an upcast from the intermediate
type to the target type (a \emph{threesome}).

@@

The indexing of tables appears 130448 times:
86\% of them are for reading a table field while
14\% of them are for writing into a table field.
We can classify the indexing operations that are reads as follows:
89\% of the reads use a literal string key,
4\% of the reads use a literal number key,
and less than 1\% of the reads (10 times) use a literal boolean key.
At all, 93\% of the reads use literals to index a table while
7\% of the reads use non-literal expressions to index a table.
It is worth mentioning that 45\% of the reads are actually
function calls.
More precisely, 25\% of the reads use literals to call a function,
20\% of the reads use literals to call a method, that is,
a function call that uses the colon syntactic sugar, 
and less than 1\% of the reads (195 times) use non-literal expressions
to call a function.
We can also classify the indexing operations that are writes as follows: 
69\% of the writes use a literal string key,
2\% of the writes use a literal number key,
and less than 1\% of the writes (1 time) uses a literal boolean key.
At all, 71\% of the writes use literals to index a table while
29\% of the writes use non-literal expressions to index a table.

We also measured how many files have code that iterates over tables to
observe how frequently iteration is used.
We observed that 23\% of the files have code that iterates over keys
of any value, that is, the call to \texttt{pairs} appears at least
once in these files (the median is twice per file);
21\% of the files have code that iterates over integer keys, that is,
the call to \texttt{ipairs} appears at least once in these files
(the median is also twice per file);
and 10\% of the files have code that use the numeric \texttt{for}
along with the length operator (the median is once per file).

### 2015-05-06 doc/thesis_puc/system.tex
Fixing typos

The subtyping rules for literal types and base types include the rules
for defining that literal types are subtypes of their respective base types,
and that $\Integer$ is a subtype of $\Number$:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-INT1}\\
\senv \vdash {\it int} \subtype \Integer
\end{array}
\;
\begin{array}{c}
\mylabel{S-INT2}\\
\senv \vdash {\it int} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-FLOAT}\\
\senv \vdash {\it float} \subtype \Number
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash \Integer \subtype \Number
\end{array}
\end{array}
\]

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rlr}
s ::= & & \textsc{statements:}\\
& \;\; \mathbf{skip} \;
| \; s_{1} \; ; \; s_{2} & \textit{skip and sequence}\\
& | \; \overline{l} = el & \textit{multiple assignment}\\
& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \;
| \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
& | \; \mathbf{return} \; el & \textit{return} \\
& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
e ::= & & \textsc{expressions:}\\
& \;\; \mathbf{nil} \;
| \; k & \textit{nil and other literals}\\
& | \; id & \textit{variable access}\\
& | \; e_{1}[e_{2}] & \textit{table access}\\
& | \; {<}T{>} \; id & \textit{type coercion}\\
& | \; f & \textit{function declaration}\\
& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
& | \; e_{1} + e_{2} \;
| \; e_{1} \; {..} \; e_{2} \;
| \; e_{1} == e_{2} \;
| \; e_{1} < e_{2} & \textit{binary operations}\\
& | \; e_{1} \;\&\; e_{2} \;
| \; e_{1} \; \mathbf{and} \; e_{2} \;
| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
& | \; \mathbf{not} \; e \;
| \; \# \; e & \textit{unary operations} \\
& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
l ::= & & \textsc{left-hand values:}\\
& \;\; id_{l} & \textit{variable assignment}\\
& | \; e_{1}[e_{2}] & \textit{table assignment}\\
& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
k ::= & & \textsc{literal constants:}\\
& \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} & \\
el ::= & & \textsc{expression lists:}\\
& \;\; \overline{e} \; | \;
\overline{e}, me & \\
me ::= & & \textsc{multiple results:}\\
& \;\; a & \textit{application}\\
& | \; {...} & \textit{vararg expression}\\
a ::= & & \textsc{applications:}\\
& \;\; e(el) & \textit{function application}\\
& | \; e{:}n(el) & \textit{method application}\\
f ::= & & \textsc{function declarations:}\\
& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
pl ::= & & \textsc{parameter lists:}\\
& \;\; \overline{id{:}T} \; | \;
\overline{id{:}T},{...}{:}T & \\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua}
\label{fig:syntax}
\end{figure}

@@

Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $T{*}$.
For instance, rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
the case where an expression list is just an expression that may produce
multiple values.
The former rule includes the type $\Nil{*}$, but the latter rule does not:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
\end{array}
\end{array}
\]

@@

Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

>>>>

The subtyping rules for literal types and base types include the rules
for defining that literal types are subtypes of their respective base types,
and that $\Integer$ is a subtype of $\Number$:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-INT1}\\
\senv \vdash {\it int} \subtype \Integer
\end{array}
\;
\begin{array}{c}
\mylabel{S-INT2}\\
\senv \vdash {\it int} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-FLOAT}\\
\senv \vdash {\it float} \subtype \Number
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-INTEGER}\\
\senv \vdash \Integer \subtype \Number
\end{array}
\end{array}
\]

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rlr}
s ::= & & \textsc{statements:}\\
& \;\; \mathbf{skip} & \textit{skip}\\
& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
& | \; \overline{l} = el & \textit{multiple assignment}\\
& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
& | \; \mathbf{return} \; el & \textit{return} \\
& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
e ::= & & \textsc{expressions:}\\
& \;\; \mathbf{nil} & \textit{nil}\\
& | \; k & \textit{other literals}\\
& | \; id & \textit{variable access}\\
& | \; e_{1}[e_{2}] & \textit{table access}\\
& | \; {<}T{>} \; id & \textit{type coercion}\\
& | \; f & \textit{function declaration}\\
& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
& | \; e_{1} + e_{2} \;
| \; e_{1} \; {..} \; e_{2} \;
| \; e_{1} == e_{2} \;
| \; e_{1} < e_{2} & \textit{binary operations}\\
& | \; e_{1} \;\&\; e_{2} \;
| \; e_{1} \; \mathbf{and} \; e_{2} \;
| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
& | \; \mathbf{not} \; e \;
| \; \# \; e & \textit{unary operations} \\
& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
l ::= & & \textsc{left-hand values:}\\
& \;\; id_{l} & \textit{variable assignment}\\
& | \; e_{1}[e_{2}] & \textit{table assignment}\\
& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
k ::= & & \textsc{literal constants:}\\
& \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} & \\
el ::= & & \textsc{expression lists:}\\
& \;\; \overline{e} \; | \;
\overline{e}, me & \\
me ::= & & \textsc{multiple results:}\\
& \;\; a & \textit{application}\\
& | \; {...} & \textit{vararg expression}\\
a ::= & & \textsc{applications:}\\
& \;\; e(el) & \textit{function application}\\
& | \; e{:}n(el) & \textit{method application}\\
f ::= & & \textsc{function declarations:}\\
& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
pl ::= & & \textsc{parameter lists:}\\
& \;\; \overline{id{:}T} \; | \;
\overline{id{:}T},{...}{:}T & \\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua}
\label{fig:syntax}
\end{figure}

@@

Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $T{*}$.
For instance, rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
the case where an expression list is just an expression that may produce
multiple values.
The former rule includes the type $\Nil{*}$, but the latter rule does not:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{1}
\end{array}
\end{array}
\]

@@

Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]
